---
title: "Lab2"
author: "Isa Berdecio"
date: "2026-09-16"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Data Import

```{r}
url <- paste0("https://www.ncbi.nlm.nih.gov/geo/download/",
              "?acc=GSE52778&format=file&file=GSE52778_All_Sample_FPKM_Matrix.txt.gz")

# Download the file
download.file(
  url,
  "GSE52778_All_Sample_FPKM_Matrix.txt.gz",
  mode = "wb"
)

# Read the compressed tab-delimited file
fpkm <- read.table(
  gzfile("GSE52778_All_Sample_FPKM_Matrix.txt.gz"),
  header = TRUE,
  sep = "",
  check.names = FALSE
)

```

# Normalize Data

```{r}

#log transform to standardize for comparison of all 4 treatments (dexamethasone, 
#albuterol, dexamethasone+albuterol or  left untreated)


#columns of interest
treatments <- fpkm[, c(
  "Untreated_FPKM",
  "Dex_FPKM",
  "Alb_FPKM",
  "Alb_Dex_FPKM"
)]


#log transformation
log_data <- log2(treatments + 1)


```

# How much does each gene change across treatments?

```{r}

#how much each gene changes across the 4 treatments
gene_range <- apply(log_data, 1, function(x) max(x) - min(x))

#top 10 genes differences
top10_index <- order(gene_range, decreasing = TRUE)[1:10]

top10 <- log_data[top10_index, ]
gene_names <- fpkm$gene_short_name[top10_index]
```

# Plot changes
```{r}
barplot(
  t(as.matrix(top10)),
  beside = TRUE,
  names.arg = gene_names,
  las = 2,
  ylab = "log2(FPKM + 1)",
  main = "Genes with Largest Expression Differences Across Treatments",
  legend.text = c(
    "Untreated",
    "Dexamethasone",
    "Albuterol",
    "Dex + Albuterol"
  )
)
```

# Interpretation

Several genes showed distinct differences in expression across treatments. Some genes like SNORD54 werer highly expressed in untreated cells only, while other like FTL and SNORD25 showed stronger expression in the 3 applied treatment groups. This pattern suggests that treatments have distinct effects on gene expression.

# Comparison of Python vs. R

The R was faster for me. I am more familiar with Markdown and Knitting, and am more well versed in data visualization using R. Additionally, it easier to preview the dataset in RStudio instead of Jupyter, making it more straightforward to see what was needed for transformation. 

